Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-06 main.Rmd
data structure section content cut

@@

## Synergy of the cyclic granularities {#sec:synergy}

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). While considering data sets of the form $<C_1, C_2, v>$, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ data subsets. We will discuss few cases, where one or more of these $nm$ sets will be empty, the implication of which we will discuss in Section \ref{sec:vis-interaction}.

@@

<!-- Let's take a specific example, where $C_1$ maps row numbers to Day-of-Month and $C_2$ maps row numbers to Week-of-Month. Here $C_1$ can take 31 values while $C_2$ can take 5 values. There will be $31\times 5=155$ sets $S_{ij}$ corresponding to the possible combinations of WOM and DOM. Many of these are empty. For example $S_{1,5}$, $S_{21,2}$, etc. In fact, most of these 155 sets will be empty, making the combination of $C_1$ and $C_2$ in a graph unhelpful. These are structurally empty sets in that it is impossible for them to have any observations. -->

<!-- Another example could be where $C_1$ maps row numbers to Day-of-Week and $C_2$ maps row numbers to Month-of-Year. Here $C_1$ can take 7 values while $C_2$ can take 12 values. So there are $12\times7=84$ sets $S_{ij}$ corresponding to the possible combinations of DOW and MOY. All of these are non-empty because every DOW can occur in every month. So graphics involving $C_1$ and $C_2$ are potentially useful. -->

<!-- What  is interaction and some examples why it is important to study interaction -->

<!-- \autoref{fig:allFig} (c) shows box plot of electricity consumption of Victoria from 2012 to 2014 across days of the year by the 1st, 15th, 29th and 31st days of the month. -->

<!-- Letter value plots convey detailed information about tails of the distribution and outliers are unexpected observations rather than extreme observations. M, F, E, D and C represents 50%, 25%, 12.5%, 6.25% and 3.13% of the tail area respectively. Some observations that can be made from these letter-value plots include a) Right tails for most days in January lying above the median is more extended than the left tails contrary to days in April which has longer left tails. b) Days in mid January and February are characterized by high variation in consumption level, but days in January have longer right tails. c) Last five days in December (Christmas holidays) shows low variation in consumption with longer left tails implying people typically consume less electricity. We can conclude that Month-of-Year and Day-of-Month are harmonies. -->

<!-- \autoref{fig:allFig} (c) shows box plot of electricity consumption of Victoria from 2012 to 2014 across days of the year by the 1st, 15th, 29th and 31st days of the month. The box plot is a a very compact distributional summary, displaying median, quartile boundaries, hinges, whiskers and outliers. All facets do not contain data across same x-axis levels leading to difficulty in comparison across facets. Hence, Day-of-Month and Day-of-Year are clashes. -->

<!-- Take another example \autoref{fig:allFig} (d) showing violin plot across days of the month faceted by week of the month. Violin plots are a combination of box plot and density plot. Here, the first week of the month correspond to certain days of the month which are different for different week of the month. These kinds of graphics can hinder our comprehension across different weeks of the month and can be categorized as a incompatible combination to plot. -->

<!-- In \autoref{fig:allFig} (e), variations across week of the year conditional on week of the month can be observed through a ridge plot. The y-axis represents week of the year and the x-axis represents electricity consumption. Ridge plots are density plots all aligned to the same horizontal scale and presented with a slight overlap. They are designed to bring out changes in distributions over time. The data is distributed unequally for different levels of weeks of the year for different facets, which hinders comparison across facets. So, Week-of-Month and Week-of-Year are clashes. -->

<!-- \autoref{fig:allFig} (f) shows decile plots of consumption across  day of the year and month of the year. Decile plots are useful in displaying distribution through deciles without much clutter. This plot, however, seem ineffective for comparison as levels of x-axis for which observed data is plotted are completely disjoint across facets. So Month-of-Year and Day-of-Year are clashes as well. -->

<!-- In \autoref{fig:allFig} (c), we have empty combinations when we plot observations with "Day-of-Month" and "Day-of-Year". Here, the 1st day of the month can never correspond to 2nd, 3rd or 4th day of the year. Hence,  "Day-of-Month" and "Day-of-Year" are clashes due to the way they map to calendar. -->

<!-- In \autoref{fig:allFig} (a), we will not have any empty combinations because every DoM can occur in all MoY. Here, mapping from days to months is irregular in the sense that one month can consist of 28, 29, 30 or 31 days. There is no denying that the 29th, 30th or 31st day of the month needs to be analysed with caution due to the irregular mapping, however, the first 28 days of the month will occur in all months of the year. -->

@@

<!-- Before exploring the behavior of a "dependent variable" across time granularities, -->
<!-- it is important to know how these granularities interact with each other. If two time granulaties of interest interact, the relationship between each of the interacting variables and the dependent variable will depend on the value of the other interacting variable.  -->

@@

<!-- When we plot some observations with two circular or aperiodic granularities used as aesthetics or facets, problems arise when we have empty combinations. It is important to check how these two granularities interact with each other, before we expect to gain insights by plotting them together.  -->

<!-- Suppose we have two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1,A_2,A_3,\dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1,B_2,B_3,\dots, B_m$\}.  That is, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = i$ and $C_2(s) = j$. Since, $C_1$ has n levels and $C_2$ has m levels there will be $nm$ such sets $S_{ij}$.  -->

<!-- Now, many situations can lead to some of these sets being empty. Let us discuss the following cases, where one or more of these sets can be empty. -->

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy. These are called "structurally" empty combinations. Let us take a specific example, where $C_1$ be day-of-month with 31 levels and $C_2$ be week-of-month with 5 levels. There will be $31\times 5=155$ sets $S_{ij}$ corresponding to possible combinations of $C_1$ and $C_2$. Many of these like $S_{1,5}$, $S_{21,2}$ are empty. This is also intuitive since the first day of the month can never correspond to fifth week of the month.

@@

An example when there will be no empty combinations could be where $C_1$ and $C_2$ maps index set to day-of-week and month-of-year respectively. Here $C_1$ can have 7 levels while $C_2$ can have 12 levels. So there are $12\times7=84$ sets $S_{ij}$. All of these are non-empty because every day-of-week can occur in every month.

A pair of cyclic granularities which lead to structurally, event-driven or build-based empty-combinations are referred to as to as **clashes**. And the ones that do not lead to any missing combinations are referred as **harmonies**.

Summarizing properties of the distribution of the measured variable can give valuable indication of features like central tendency, skewness, multimodality, tail behavior or variation. However, we should have the consider the effect of number of observations on the summarization. Suppose we have $T$ observations, and two cyclic granularities $C_1$ with $n$ categories and $C_2$ with $m$ categories. Each element of $C_1$ occurs approximately $T/n$ times while each element of $C_2$ occurs approximately $T/m$ times. There are no empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(mn)$. If we require at least $k$ observations to create a meaningful panel, then provided $T \ge mnk$, the visualization will be acceptable. The value of $k$ will depend on the statistical transformation we are producing. For computing deciles, $k=10$ may be acceptable, but for density estimates, we would need $k \ge 30$. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.

>>>>

## Data structure for exploring two cyclic granularities

@@

## Harmonies and Clashes {#sec:synergy}

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). While considering data sets of the form $<C_1, C_2, v>$, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ data subsets. 
Now, some situations can lead to some of these sets being empty. We will discuss few cases, where one or more of these $nm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

\begin{definition}[Clash]\label{def:definition}
A pair of cyclic granularities which contains structurally, event-driven or build-based empty combinations of its categories.
\end{definition}

\begin{definition}[Harmony]\label{def:definition}
A pair of cyclic granularities that do not contain any missing combinations of its categories.
\end{definition}

@@

@@

@@

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy. These are called "structurally" empty combinations. Let us take a specific example, where $C_1$ be day-of-month with 31 levels and $C_2$ be week-of-month with 5 levels. There will be $31\times 5=155$ sets $S_{ij}$ corresponding to possible combinations of $C_1$ and $C_2$. Many of these like $S_{1,5}$, $S_{21,2}$ are empty. This is also intuitive since the first day of the month can never correspond to fifth week of the month. Hence the pair (day-of-month, week-of-month) is a clash.

@@

An example when there will be no empty combinations could be where $C_1$ and $C_2$ maps index set to day-of-week and month-of-year respectively. Here $C_1$ can have 7 levels while $C_2$ can have 12 levels. So there are $12\times7=84$ sets $S_{ij}$. All of these are non-empty because every day-of-week can occur in every month. Hence, the pair (day-of-week, month-of-year) is a harmony.

Summarizing properties of the distribution of the measured variable can give valuable indication of features like central tendency, skewness, multimodality, tail behavior or variation. However, we should have the consider the effect of number of observations on the summarization even for harmonies. Suppose we have $T$ observations, and two cyclic granularities $C_1$ with $n$ categories and $C_2$ with $m$ categories. Each element of $C_1$ occurs approximately $T/n$ times while each element of $C_2$ occurs approximately $T/m$ times. There are no empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(mn)$. If we require at least $k$ observations to create a meaningful panel, then provided $T \ge mnk$, the visualization will be acceptable. The value of $k$ will depend on the statistical transformation we are producing. For computing deciles, $k=10$ may be acceptable, but for density estimates, we would need $k \ge 30$. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.
